High-level cognition is supported by at least second order dynamic correlations in neural activity patterns

### 2019-09-03 paper/boneyard.tex
backing up progress on results section: decoding results
finished pass through results section

>>>>

% finally, we sought to identify which specific brain areas were
% involved in supporting cognition through high-order interactions.
% we used the eigenvector centrality results, which preserves the
% original feature space when characterizing high-order dynamic
% correlations.  we found that first and second-order correlations
% reflected auditory and speech processing in all of the story
% listening conditions.  during intact story listening, third order
% correlations reflected integration with visual areas, and fourth
% order correlations recruited high-level cognition regions.  however,
% during listening to temporally scrambled stories, these higher-order
% correlations did not involve visual areas or high-order cognitive
% control regions.  during rest we found a much different set of
% patterns.  first order correlations involved counting networks; second-order
% correlations recruited visual networks; third-order correlations
% recruited working memory networks; fourth-order correlations
% recruited episodic memory areas.

\subsubsection*{Reverse inference}
The dynamic patterns we examine comprise high-dimensional correlation
patterns at each timepoint.  To help interpret the resulting patterns
in the context of other studies, we created summary maps by computing
the across-timepoint average pairwise correlations at each order of
analysis (first order, second order, etc., up to fifteenth order
correlations).  We selected the 10 strongest (absolute value)
correlations at each order.  Each correlation is between the dynamic
activity patterns (or patterns of dynamic high-order correlations)
measured at two RBF nodes (see \textit{Hierarchical Topographic Factor
  Analysis}).  Therefore, the 10 strongest correlations involved up to
20 RBF nodes.  Each RBF defines a spatial function whose activations
range from 0 to 1.  We thresholded each RBF at 0.999 to construct a
map of spherical components that denoted the endpoints of the 10
strongest correlations.  We then carried out a meta analysis using
Neurosynth~\citep{RubiEtal17} to identify the 10 terms most commonly
associated with the given map.  This resulted in a set of 10 terms
associated with the average dynamic correlation patterns at each
order.

%%%%
For each order $n \in \{1, 2, 3, 4}$
we used the selected kernel to compute the $(n-1)^\mathrm{th}$-order
DISFC timeseries across all participants (for $n = 1$ we simply
averaged the ``order 0'' activity timeseries across participants).  We
then computed the average $n^\mathrm{th}$-order correlations (over
time) by correlating the columns of the $(n-1)^\mathrm{th}$-order
DISFC timeseries.  This yielded a single $K$ by $K$
$n^\mathrm{th}$-order correlation matrix.  Separately for each
experimental condition and order of correlations, we identified the
endpoints of the 10 strongest (absolute value) correlations
(Fig.~\ref{fig:neurosynth}).

We used this kernel to compute 
%%%%

% we used temporal decoders to identify cognitivly relevant neural
% patterns in fMRI data (see forward inference section of methods).
% we explored two approaches to effiently computing high-order
% correlations: PCA (dimensionality reduction) and eigenvector
% centrality (graph measure).  both approaches yielded qualitatively
% similar results for the story listening conditions: intact story was
% best decoded using high-order correlations, whereas scrambled
% stories were best decoded using lower-order correlations.  both
% approaches also yielded slightly above-chance decoding of resting state times
% (perhaps picking up on decreasing attention and/or increasing
% mind-wandering as the resting state scan progressed).  The PCA-based
% approach achieved the highest decoding accuracy using
% non-correlational (activation-based) data, whereas the eigenvector
% centrality approach identified second-order correlations as being
% most informative about resting state times.

### 2019-09-03 paper/main.tex
finished pass through results section

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figs/synthetic_data.pdf}
  \caption{\textbf{Recovering known dynamic correlations from
      synthetic data.}  Each panel displays the average correlations
    between the vectorized upper triangles of the recovered
    correlation matrix at each timepoint and either the true
    underlying correlation at each timepoint or a reference
    correlation matrix.  (The averages are taken across 10 different
    randomly generated synthetic datasets of the given category.)
    Error ribbons denote 95\% confidence intervals (taken across
    datasets). Different colors denote different kernel shapes,
    whereas the shading within each color type denotes kernel widths.
    For a complete description of each synthetic dataset, see
    \textit{Synthetic data}.  \textbf{a. Constant correlations.} These
    datasets have a stable (unchanging) underlying correlation matrix.
    \textbf{b. Random correlations.}  These datasets are generated
    using a new independently drawn correlation matrix at each new
    timepoint.  \textbf{c. Ramping correlations.}  These datasets are
    generated by smoothly varying the underlying correlations between
    the randomly drawn correlation matrices at first and last
    timepoints.  The left panel displays the correlations between the
    recovered dynamic correlations and the underlying ground truth
    correlations.  The middle panel compares the recovered
    correlations with the \textit{first} timepoint's correlation
    matrix.  The right panel compares the recovered correlations with
    the \textit{last} timepoint's correlation matrix.
    \textbf{d. Event-based correlations.}  These datasets are each
    generated using five randomly drawn correlation matrices that each
    remain stable for a fifth of the total timecourse.  The left panel
    displays the correlations between the recovered dynamic
    correlations and the underlying ground truth correlations.  The
    right panels compare the recovered correlations with the
    correlation matrices unique to each event.}
  \label{fig:synthetic}
\end{figure}

@@

\begin{figure}[hp]
  \centering
  \includegraphics[width=\textwidth]{figs/decode_level.pdf}
  \caption{\textbf{Across-participant decoding accuracy varies with
      correlation order and cognitive engagement.}
    \textbf{a. Decoding accuracy as a function of order: PCA.}
    \textit{Order} ($x$-axis) refers to the maximum order of dynamic
    correlations that were available to the classifiers (see
    \textit{Feature weighting and testing}).  The reported
    across-participant decoding accuracies are averaged over all
    kernel shapes and widths (see \textit{Identifying robust decoding
      results}).  The $y$-values are displayed relative to chance
    accuracy (intact: $\frac{1}{300}$; paragraph: $\frac{1}{272}$;
    word: $\frac{1}{300}$; rest: $\frac{1}{400}$).  The error bars
    denote 95\% confidence intervals across cross-validation folds
    (i.e., random assignments of participants to the training and test
    sets).  The colors denote the experimental condition.  Arrows
    denote sets of features that yielded reliably higher (upwards
    facing) or lower (downward facing) decoding accuracy than the mean
    of all other features (via a two-tailed test, thresholded at
    $p < 0.05$).  The circled values represent the maximum decoding
    accuracy within each experimental condition.
    \textbf{b. Normalized decoding accuracy as a function of order:
      PCA.}  This panel displays the same results as Panel a, but here
    each curve has been normalized to have a maximum value of 1 and
    and a minimum value of 0 (including the upper and lower bounds of
    the respective 95\% confidence intervals).  Panels a and b used
    PCA to project each high-dimensional pattern of dynamic
    correlations onto a lower-dimensional space.  \textbf{c.  Decoding
      accuracy as a function of order: eigenvector centrality.} This
    panel is in the same format as Panel a, but here eigenvector
    centrality has been used to project the high-dimensional patterns
    of dynamic correlations onto a lower-dimensional space.
    \textbf{d. Noramlized decoding accuracy as a function of order:
      eigenvector centrality.} This panel is in the same format as
    Panel b, but here eigenvector centrality has been used to project
    the high-dimensional patterns of dynamic correlations onto a
    lower-dimensional space.}
  \label{fig:decoding}
\end{figure}

@@

@@

% finally, we sought to identify which specific brain areas were
% involved in supporting cognition through high-order interactions.
% we used the eigenvector centrality results, which preserves the
% original feature space when characterizing high-order dynamic
% correlations.  we found that first and second-order correlations
% reflected auditory and speech processing in all of the story
% listening conditions.  during intact story listening, third order
% correlations reflected integration with visual areas, and fourth
% order correlations recruited high-level cognition regions.  however,
% during listening to temporally scrambled stories, these higher-order
% correlations did not involve visual areas or high-order cognitive
% control regions.  during rest we found a much different set of
% patterns.  first order correlations involved counting networks; second-order
% correlations recruited visual networks; third-order correlations
% recruited working memory networks; fourth-order correlations
% recruited episodic memory areas.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/direction_of_field.pdf}
  \caption{\textbf{Direction of the field (adapted from~\citep{Turk13}).} The evolution of fMRI analyses started with
    univariate activation, which refers to
    the average amplitude of BOLD activity evoked by events of an
    experimental condition. Next, multivariate classifiers are trained
    on patterns of activation across voxels to decode distributed
    representations for specific events. The next, resting
    connectivity, is the temporal correlation of one or more seed
    regions with the remainder of the brain during rest. Additionally,
    task-based connectivity examines how these correlations differ by
    cognitive state. Following this increasing trajectory of
    increasing complexity, full connectivity considers all pairwise
    correlations in the brain, most commonly at rest.  Next, dynamic
    full connectivity considers how full connectivity changes over
    time. Continuing this line of reasoning, we expect higher-order network dynamics might provide even richer insights into the neural basis of cognition.}
  \label{fig:direction_of_field}
\end{figure}

>>>>

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figs/synthetic_data}
  \caption{\textbf{Recovering known dynamic correlations from
      synthetic data.}  Each panel displays the average correlations
    between the vectorized upper triangles of the recovered
    correlation matrix at each timepoint and either the true
    underlying correlation at each timepoint or a reference
    correlation matrix.  (The averages are taken across 10 different
    randomly generated synthetic datasets of the given category.)
    Error ribbons denote 95\% confidence intervals (taken across
    datasets). Different colors denote different kernel shapes,
    whereas the shading within each color type denotes kernel widths.
    For a complete description of each synthetic dataset, see
    \textit{Synthetic data}.  \textbf{a. Constant correlations.} These
    datasets have a stable (unchanging) underlying correlation matrix.
    \textbf{b. Random correlations.}  These datasets are generated
    using a new independently drawn correlation matrix at each new
    timepoint.  \textbf{c. Ramping correlations.}  These datasets are
    generated by smoothly varying the underlying correlations between
    the randomly drawn correlation matrices at first and last
    timepoints.  The left panel displays the correlations between the
    recovered dynamic correlations and the underlying ground truth
    correlations.  The middle panel compares the recovered
    correlations with the \textit{first} timepoint's correlation
    matrix.  The right panel compares the recovered correlations with
    the \textit{last} timepoint's correlation matrix.
    \textbf{d. Event-based correlations.}  These datasets are each
    generated using five randomly drawn correlation matrices that each
    remain stable for a fifth of the total timecourse.  The left panel
    displays the correlations between the recovered dynamic
    correlations and the underlying ground truth correlations.  The
    right panels compare the recovered correlations with the
    correlation matrices unique to each event.}
  \label{fig:synthetic}
\end{figure}

@@

\begin{figure}[hp]
  \centering
  \includegraphics[width=\textwidth]{figs/decode_level}
  \caption{\textbf{Across-participant decoding accuracy varies with
      correlation order and cognitive engagement.}
    \textbf{a. Decoding accuracy as a function of order: PCA.}
    \textit{Order} ($x$-axis) refers to the maximum order of dynamic
    correlations that were available to the classifiers (see
    \textit{Feature weighting and testing}).  The reported
    across-participant decoding accuracies are averaged over all
    kernel shapes and widths (see \textit{Identifying robust decoding
      results}).  The $y$-values are displayed relative to chance
    accuracy (intact: $\frac{1}{300}$; paragraph: $\frac{1}{272}$;
    word: $\frac{1}{300}$; rest: $\frac{1}{400}$).  The error bars
    denote 95\% confidence intervals across cross-validation folds
    (i.e., random assignments of participants to the training and test
    sets).  The colors denote the experimental condition.  Arrows
    denote sets of features that yielded reliably higher (upwards
    facing) or lower (downward facing) decoding accuracy than the mean
    of all other features (via a two-tailed test, thresholded at
    $p < 0.05$).  The circled values represent the maximum decoding
    accuracy within each experimental condition.
    \textbf{b. Normalized decoding accuracy as a function of order:
      PCA.}  This panel displays the same results as Panel a, but here
    each curve has been normalized to have a maximum value of 1 and
    and a minimum value of 0 (including the upper and lower bounds of
    the respective 95\% confidence intervals).  Panels a and b used
    PCA to project each high-dimensional pattern of dynamic
    correlations onto a lower-dimensional space.  \textbf{c.  Decoding
      accuracy as a function of order: eigenvector centrality.} This
    panel is in the same format as Panel a, but here eigenvector
    centrality has been used to project the high-dimensional patterns
    of dynamic correlations onto a lower-dimensional space.
    \textbf{d. Noramlized decoding accuracy as a function of order:
      eigenvector centrality.} This panel is in the same format as
    Panel b, but here eigenvector centrality has been used to project
    the high-dimensional patterns of dynamic correlations onto a
    lower-dimensional space.}
  \label{fig:decoding}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{figs/most_abs}
  \caption{\textbf{Top terms associated with the endpoints of the
      strongest correlations.}  Each color corresponds to one order of
    correlations, averaged across participants and time. The inflated
    brain plots display the locations of the endpoints of the 10
    strongest (absolute value) correlations at each order, projected
    onto the cortical surface~\citep{CombEtal19}.  The lists of terms
    on the right display the top five Neurosynth
    terms~\citep{RubiEtal17} decoded from the corresponding brain maps
    for each order.  Each row displays data from a different
    experimental condition.  Additional maps and their corresponding
    Neurosynth terms may be found in the \textit{Supplementary
      materials} (intact: Figure S1; paragraph: Figure S2; word:
    Figure S3; rest: Figure S4).}
  \label{fig:neurosynth}
\end{figure}

Having established that patterns of high-order correlations are
informative to decoders, we next wondered which specific networks of
brain regions contributed most to these patterns.  As a representative
example, we selected the kernel parameters that yielded decoding
accuracies that best matched the average accuracies across all of the
kernel parameters we examined.  Using Figure~\ref{decoding}c as a
template, the best-matching kernel was a Laplace kernel with a width
of 50 (Fig.~\ref{fig:kernels}d).  We used this kernel to compute a
single $K$ by $K$ $n^\mathrm{th}$-order correlation matrix for each
experimental condition, along with the Neurosynth~\citep{RubiEtal17}
terms most highly associated with each of these matrices
(Fig.~\ref{fig:neurosynth}; see \textit{Reverse inference}).

For all of the story listening conditions (intact, paragraph, and
word), we found that first- and second-order correlations were most
strongly associated with auditory and speech processing.  During
intact story listening, third-order correlations reflected integration
with visual areas, and fourth-order correlations reflected integration
with areas associated with high-level cognition and cognitive control,
such as the ventrolateral prefrontal cortex.  However, during
listening to temporally scrambled stories, these higher-order
correlations instead involved interactions with additional regions
associated with speech and semantic processing.  By contrast, we found
a much different set of patterns in the resting state data.
First-order resting state correlations were most strongly associated
with regions involved in counting and numerical understanding.
Second-order resting state correlations were strongest in visual
areas; third-order correlations were strongest in task-positive areas;
and fourth-order correlations were strongest in regions associated
with autobiographical and episodic memory.  We carried out analogous
analyses to create maps (and decode the top associated Neurosynth
terms) for up to fifteenth-order correlations (Figs. S1, S2, S3, and
S4).  Of note, examining fifteenth-order correlations between 700
nodes using conventional methods would have required storing roughly
$\frac{700^{2 \times 15}}{2} \approx 1.13 \times 10^{85}$ floating point
numbers-- assuming single-precision (32 bits each), this would require
roughly 32 times as many bits as there are molecules in the known
universe!  Although these fifteenth-order correlations do appear
(visually) to have some well-formed structure, we provide this latter
example primarily as a demonstration of the efficiency and scalability
of our approach.

@@

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/direction_of_field}
  \caption{\textbf{Direction of the field (adapted from~\citep{Turk13}).} The evolution of fMRI analyses started with
    univariate activation, which refers to
    the average amplitude of BOLD activity evoked by events of an
    experimental condition. Next, multivariate classifiers are trained
    on patterns of activation across voxels to decode distributed
    representations for specific events. The next, resting
    connectivity, is the temporal correlation of one or more seed
    regions with the remainder of the brain during rest. Additionally,
    task-based connectivity examines how these correlations differ by
    cognitive state. Following this increasing trajectory of
    increasing complexity, full connectivity considers all pairwise
    correlations in the brain, most commonly at rest.  Next, dynamic
    full connectivity considers how full connectivity changes over
    time. Continuing this line of reasoning, we expect higher-order network dynamics might provide even richer insights into the neural basis of cognition.}
  \label{fig:direction_of_field}
\end{figure}
